### 2020-06-12 chapters/abstract.tex
HyperLogLog. PDIS delivery.
Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory (HTM)

Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
As such, many large companies take this into consideration and deploy data anomaly components in their systems. In this way, system administrators can proactively adjust and prevent potential service failures.

For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases. At Microsoft, their custom, internally made anomaly detection system monitors millions of metrics coming from a variety of different internal services, enabling engineers to solve live on-site issues before they scale and become bigger ones.

Many real-time stream monitoring systems are static once deployed in a production environment. Engineers of said systems usually configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.

With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.

\vspace*{10mm}\noindent
\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, variety, seasonal

>>>>

%Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
%As such, many large companies take this into consideration and deploy data anomaly components in their systems. In this way, system administrators can proactively adjust and prevent potential service failures.

%For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases. At Microsoft, their custom, internally made anomaly detection system monitors millions of metrics coming from a variety of different internal services, enabling engineers to solve live on-site issues before they scale and become bigger ones.

%Many real-time stream monitoring systems are static once deployed in a production environment. Engineers of said systems usually configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.

%With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.

\vspace*{10mm}\noindent
\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, seasonal

### 2020-06-12 chapters/sota.tex
Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory (HTM)

\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}

@@

\subsubsection*{Method classification}
Extra-N can take both uni or multivariate time-series as input data. However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables. In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method is of univariate nature. Last but not least, Extra-N aims to detect univariate point outliers. These point outliers are identified based on the density of the region or neighborhood they belong too. Hence, we say Extra-N is a density-based method (Figure \ref{fig:point-outlier-taxonomy}).

@@

\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
\subsubsection*{Method classification}
\subsubsection*{Applicability to our Hypothesis}

\iffalse
HTM\\
point outlier, model-based, prediction\\
black-box ML algorithm\\
explain properties\\
use what authors say in their review\\
\fi
\subsection{DeepAnT}
\subsubsection*{Method classification}
\subsubsection*{Applicability to our Hypothesis}

>>>>

\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}

@@

\subsubsection*{Method classification}
Extra-N can take both uni or multivariate time-series as input data. However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables. In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method accepts uni or multivariate time-series as input but is of univariate nature. Last but not least, Extra-N aims to detect univariate point outliers. These point outliers are identified based on the density of the region or neighborhood they belong too. Hence, we say Extra-N is a density-based univariate point outlier detection method (Figure \ref{fig:point-outlier-taxonomy}).

@@

\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
Ahmad \& Purdy, 2016, present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}. HTM has a few properties that are desirable for data stream anomaly detection scenarios. First, it is an unsupervised algorithm that requires no labels. This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent. Additionally, HTM adapts to noisy domains. In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored. Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.

HTM is an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream. In other words, if there is a sudden abrupt change in the distribution of a specific time-dependent variable, the algorithm will raise an alert. However, if that initially thought to be random spike becomes frequent, the algorithm adapts and assimilates this as "a new reality" and stops classifying it as an anomaly. Unfortunately, this property is not desirable in our system. Our stream monitoring system must report changes relatively to an initial static configuration (reference or normality period) and should not accept a new distribution as the new standard.

\subsubsection*{Method classification}
Similarly to Extra-N presented in Section \ref{alg:extran}, the Hierarchical Temporal Memory (HTM) algorithm accepts both uni and multivariate time-series as input data but does not perform a multivariate analysis. Hence, we say it accepts both uni and multivariate inputs but is univariate in nature. HTM focuses on finding point outliers, but unlike Extra-N, which is density-based, HTM is a model-based approach. Specifically, HTM is a prediction model-based approach for univariate point outlier detection.

\subsubsection*{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

\subsection{DeepAnT}
\subsubsection*{Method classification}
\subsubsection*{Applicability to our Hypothesis}
